Hydras and Cie \\ Variations on Kirby \& Paris' hydra battles and ordinal numbers in Coq

### 2021-09-23 doc/JFLA2022/paper.tex
minor corrections

\usepackage[firstpageonly=false, color={[gray]{0.5}},
   scale=2.0, text=DRAFT]{draftwatermark}
% -------------------------------
%%%% For Alectryon

@@

\author{
Pierre Castéran \inst{1}
\and
    Jérémy Damour \inst{2}
\and
Karl Palmskog \inst{2}
\and Clément Pit-Claudel \inst{2}
\and Théo Zimmermann \inst{2}
}

\institute{
Univ. Bordeaux, CNRS, Bordeaux INP, LaBRI, UMR 5800, F-33400 Talence, France \\
  \email{pierre.casteran@labri.fr}
\and
   to fill!
 }

@@

\begin{abstract}
 To do \dots
\end{abstract}

@@

Hydra games (also known as \emph{Hydra battles}) appear in an article published in 1982 by two mathematicians, 
Laurie Kirby and Jeff Paris: \emph{Accessible Independence Results for Peano Arithmetic}~\cite{KP82}.
This article describes a game between two players: Hercules and a hydra.
A short description of the game  can be found in~\cite{bauer2008, KP82, JFLA2018paper}. One can also play with
Andrej Bauer's simulator~\cite{BauerHydra}.
In a few words:
\begin{itemize}
\item A hydra is a finite tree, traditionnaly presented with the root at the bottom, the leaves of which are called \emph{heads}
  (Fig~\ref{fig:iota-example}).
\item At every round, Hercules chops off one head of the hydra. If the head is at a distance greater or equal than 2 from the root,
  then some sub-tree $h$ of the hydra is copied a certain amount $n$ of times. The number $n$ of copies and the sub-tree $h$ may depend of the considered variant of the game
  or the time elapsed since the beginning of the fight. 
\end{itemize}
   \begin{figure}[htb]
\centering
\begin{tikzpicture}[very thick, scale=0.3]
\node (foot) at (10,0) {$\bullet$};
\node (N1) at (2,2) {$\bullet$};
\node (N2) at (10,2) {$\bullet$};
\node (N22) at (7,2) {$\bullet$};
\node (N3) at (14,2) {$\bullet$};
\node (N4) at (18,2) {$\Smiley[2][green]$};
\node (N5) at (0,4) {$\bullet$};
\node (N6) at (2,5) {$\Smiley[2][green]$};
\node (N7) at (4,6) {$\Smiley[2][green]$};
\node (N88) at (7,4) {$\bullet$};
\node (N8) at (10,4) {$\bullet$};
\node (N9) at (14,6) {$\Smiley[2][green]$};
\node (N10) at (0,8) {$\Smiley[2][green]$};
\node (N11) at (10,7) {$\Smiley[2][green]$};
\node (N111) at (7,7) {$\Smiley[2][green]$};
\draw (foot) to [bend left=10] (N1);
\draw (foot) -- (N2);
\draw (foot) -- (N22);
\draw (foot) -- (N3);
\draw (foot) -- (N4);
\draw (N1) to  (N5);
\draw (N1) to   [bend left=10] (N6);
\draw (N1) to   [bend right=20] (N7);
\draw (N2) to  (N8);
\draw (N22) to  (N88);
\draw (N8) to  (N11);
\draw (N88) to  (N111);
\draw (N3) to  (N9);
\draw (N5) to  (N10);
\end{tikzpicture}
\caption{The hydra associated with the ordinal $\omega^{\omega+2}+\omega^\omega \times 2 + \omega + 1$ \label{fig:iota-example}}

\end{figure}

@@

This work is also an opportunity to 
 provide ``concrete'' examples of several formalisation and proof techniques: operational type classes, functions defined by  equations, dependently typed functions, etc. It may be also used as a library on ordinal numbers, for instance for proving termination properties.

@@

Indeed, the consideration of \emph{all} battles allowed us to
consider battles where the hydra can make an arbitrary number of copies at any time, which make easier our proof by absurd.
Unfortunately, the examples  most shown in the litterature
(see for instance \cite{KP82, bauer2008, BauerHydra}) 
assume that the hydra grows $n$ copies at step $n$ of the game, which is incompatible with our proof.

@@

 \begin{itemize}
   \item The game starts at an initial step $i\in\mathbb{N}$ (not necessariliy $0$).
   \item  The initial hydra is the graphical representation of an ordinal  $\alpha<\epsilon_0$  in Cantor normal form (See for instance Fig~\ref{fig:iota-example}). 
 \item Hercules always chops off the rightmost head of the hydra.
 \end{itemize}

@@

In order to prove formally that the length of the considered
kind of battles is not given by a primitive recursive function, weused the formalisation of primitive recursive functions, part
of Russell O'Connor's contribution on G\"{o}del first incompleteness theorem \cite{OConnor05, Goedel}.
For this purpose, and above all by consideration of the scientific interest of this contribution, we chose to host and maintain this work within \textit{Coq-Community}.

@@

  \item Presentation of the main steps of proofs, possibly interleaved with text (Figures \ref{fig:Ex42E0} and \ref{fig:Ex42-schutte}).

  \end{itemize}
\end{todo}

>>>>

%\usepackage[firstpageonly=false, color={[gray]{0.5}},
%   scale=2.0, text=DRAFT]{draftwatermark}
% -------------------------------
%%%% For Alectryon

@@

\author{
Pierre Castéran \inst{1}
\and
    Jérémy Damour \inst{2}
\and
Karl Palmskog \inst{3}
\and Clément Pit-Claudel \inst{4}
\and Théo Zimmermann \inst{5}
}

\institute{
Univ. Bordeaux, CNRS, Bordeaux INP, LaBRI, UMR 5800, F-33400 Talence, France \\
  \email{pierre.casteran@labri.fr}
\and
to fill!
\and
to fill!
\and
to fill!
\and
to fill!
 }

@@

\begin{abstract}
  \emph{Hydra-battles} is a collaborative library of commented proofs, developped under the \emph{Coq-Community} project.
  We present its evolution since a former presentation~\cite{JFLA2018paper} and how its structure is determined by other libraies and the \emph{Alectryon} literate proving tool.
\end{abstract}

@@

Hydra games (also known as \emph{Hydra battles}) appear in an article published in 1982 by two mathematicians, 
Laurie Kirby and Jeff Paris: \emph{Accessible Independence Results for Peano Arithmetic}~\cite{KP82}.
This article describes a game between two players: Hercules and a hydra.
A short description of the game  can be found in~\cite{bauer2008, KP82, JFLA2018paper}. One can also play with
Andrej Bauer's simulator~\cite{BauerHydra}.
In a few words:
\begin{itemize}
\item A hydra is a finite tree, traditionally presented with the root at the bottom, the leaves of which are called \emph{heads}
  (Fig~\ref{fig:round}).
\item At every round, Hercules chops off one head of the hydra. If the head is at a distance greater or equal than 2 from the root,
  then some sub-tree $h$ of the hydra is copied a certain amount $n$ of times. The number $n$ of copies and the sub-tree $h$ may depend of the considered variant of the game
  or the time elapsed since the beginning of the fight.
  Figure~\ref{fig:round} shows an example of such a transformation with $n=2$.
\end{itemize}

\begin{figure}[h]
  \centering
  \begin{tikzpicture}[very thick, scale=0.3]
  \node (h1) at (4,0){$\bullet$};
  \node (h2) at (4,2){$\bullet$};
  \node (h3) at (2,4){$\bullet$};
  \node (h4) at (0,6){$\bullet$};
  \node (h5) at (4,6){$\bullet$};
  \node (h6) at (6,4){$\bullet$};
  \draw (h1) -- (h2) ;
  \draw (h2) -- (h3) ;
  \draw (h2) -- (h6);
  \draw (h3) -- (h4) ;
  \draw (h3) -- (h5) ;

\node (hn1) at (14,0){$\bullet$};
\node (hn2) at (12,2) {$\bullet$};
\node (hn3) at (12,4) {$\bullet$};
\node (hn4) at (10,6){$\bullet$};
  \node (hn5) at (12,6){$\bullet$};
  \draw (hn1) -- (hn2) ;
  \draw (hn2) -- (hn3) ;
  \draw (hn3) -- (hn4) ;
  \draw (hn3) -- (hn5) ;
\node (hn2b) at (14,2) {$\bullet$};
\node (hn3b) at (14,4) {$\bullet$};
\node (hn4b) at (13.4,6){$\bullet$};
  \node (hn5b) at (15,6){$\bullet$};
  \draw (hn1) -- (hn2b) ;
  \draw (hn2b) -- (hn3b) ;
  \draw (hn3b) -- (hn4b) ;
  \draw (hn3b) -- (hn5b) ;
  \node (hn2c) at (16,2) {$\bullet$};
\node (hn3c) at (16,4) {$\bullet$};
\node (hn4c) at (17,6){$\bullet$};
  \node (hn5c) at (18.5,6){$\bullet$};
  \draw (hn1) -- (hn2c) ;
  \draw (hn2c) -- (hn3c) ;
  \draw (hn3c) -- (hn4c) ;
  \draw (hn3c) -- (hn5c) ;
\end{tikzpicture}

  \caption{Two successive states of a hydra in a battle}
  \label{fig:round}
\end{figure}

 %  \begin{figure}[htb]
% \centering
% \begin{tikzpicture}[very thick, scale=0.3]
% \node (foot) at (10,0) {$\bullet$};
% \node (N1) at (2,2) {$\bullet$};
% \node (N2) at (10,2) {$\bullet$};
% \node (N22) at (7,2) {$\bullet$};
% \node (N3) at (14,2) {$\bullet$};
% \node (N4) at (18,2) {$\Smiley[2][green]$};
% \node (N5) at (0,4) {$\bullet$};
% \node (N6) at (2,5) {$\Smiley[2][green]$};
% \node (N7) at (4,6) {$\Smiley[2][green]$};
% \node (N88) at (7,4) {$\bullet$};
% \node (N8) at (10,4) {$\bullet$};
% \node (N9) at (14,6) {$\Smiley[2][green]$};
% \node (N10) at (0,8) {$\Smiley[2][green]$};
% \node (N11) at (10,7) {$\Smiley[2][green]$};
% \node (N111) at (7,7) {$\Smiley[2][green]$};
% \draw (foot) to [bend left=10] (N1);
% \draw (foot) -- (N2);
% \draw (foot) -- (N22);
% \draw (foot) -- (N3);
% \draw (foot) -- (N4);
% \draw (N1) to  (N5);
% \draw (N1) to   [bend left=10] (N6);
% \draw (N1) to   [bend right=20] (N7);
% \draw (N2) to  (N8);
% \draw (N22) to  (N88);
% \draw (N8) to  (N11);
% \draw (N88) to  (N111);
% \draw (N3) to  (N9);
% \draw (N5) to  (N10);
% \end{tikzpicture}
% \caption{The hydra associated with the ordinal $\omega^{\omega+2}+\omega^\omega \times 2 + \omega + 1$ \label{fig:iota-example}}

% \end{figure}

@@

This work is also an opportunity to 
 provide ``concrete'' examples of formalization and proof techniques: operational type classes, functions defined by  equations, dependently typed functions, etc. It may be also used as a library on ordinal numbers, for instance for proving termination properties.

@@

Indeed, the consideration of \emph{all} battles allowed us to
consider battles where the hydra can make an arbitrary number of copies at any time, which made artificially easier our proof by absurd.
Unfortunately, the examples  most shown in the litterature
(see for instance \cite{KP82, bauer2008, BauerHydra}) 
assume that the hydra grows $n$ copies at step $n$ of the game, which is incompatible with our proof.

@@

 \begin{itemize}
   \item The game starts at an initial step $i\in\mathbb{N}$ (not necessarily $0$).
   \item  The hydra is always the the graphical representation of some ordinal strictly below $\epsilon_0$ in Cantor normal
     form. For instance, Fig~\ref{fig:round}) shows the hydras respectively associated with  $\omega^{\omega^2+1}$ and $\omega^{\omega^2}\times 3$.

 \item Hercules always chops off the rightmost head of the hydra.
 \end{itemize}

@@

In order to prove formally that the length of the considered
kind of battles is not given by a primitive recursive function, weused the formalization of primitive recursive functions, part
of Russell O'Connor's contribution on G\"{o}del first incompleteness theorem \cite{OConnor05, Goedel}.
For this purpose, and above all by consideration of the scientific interest of this contribution, we chose to host and maintain this work within \textit{Coq-Community}.

@@

  \item Presentation of the main steps of proofs, possibly interleaved with text.

  \end{itemize}

   Figures~\ref{fig:Hprime} (code snippet), \ref{fig:Ex42E0} and \ref{fig:Ex42-schutte} have been generated with Alectryon.
\end{todo}
